Machine learning topological phases in real space

### 2020-05-05 paper/main.tex
Preparing to run simulations with fewer features

\begin{abstract}
We develop a supervised machine learning algorithm that is able to learn topological phases for finite condensed matter systems in real lattice space. The algorithm employs diagonalization in real space together with any supervised learning algorithm to learn topological phases through an eigenvector-ensembling procedure. We combine our algorithm with decision trees to successfully recover topological phase diagrams of Su-Schrieffer-Heeger (SSH) models from lattice data in real space and show how the Shannon information entropy of ensembles of lattice eigenvectors can be used to retrieve a signal detailing how topological information is distributed along the lattice. The discovery of Shannon entropy signals associated with topological phase transitions from the analysis of data from several thousand SSH systems illustrates how model explainability in machine learning can advance the research of exotic quantum materials with properties that may power future technological applications such as quantum computing. 
\end{abstract}
\begin{document}
\nocite{*}

@@

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learn the desired pattern in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the whole system emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in condensed matter systems is presented in the literature.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{ssh1.png}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{ssh2.png}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiments 1 and 2 as seen from parameter space. The accuracy statistics presented in the following subsections and probability heatmaps shown in figure \ref{heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. The probability heatmaps shown in figure \ref{heatmaps} faithfully portray the phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{fig1a-6561-100-train_val_test_0.png}}\quad
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{fig1a-6561-100-train_winding_labels_0.png}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{fig1a-6561-100-prediction_grid_0.png}}
\caption{Visualization of a single iteration of experiment 1 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase prediction grid learned from real space.}
\label{figexp1_exp}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{fig1b-6561-100-train_val_test_0.png}}\quad
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{fig1b-6561-100-train_winding_labels_0.png}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{fig1b-6561-100-prediction_grid_0.png}}
\caption{Visualization of a single iteration of experiment 2 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase prediction grid learned from real space.}
\label{figexp2_exp}
\end{figure}

@@

The (normalized) relevance of a feature is given by how much it reduces a loss function (e.g. information entropy or Gini impurity \cite{friedman2001elements}). By averaging normalized relevances as measured by reduction in the information entropy of ensembles of real space eigenvectors across $n_{exp}$ = 100 iterations of both experiment 1 and experiment 2 we recovered Shannon entropy signals that reveal which lattice sites were consistently more relevant in learning topological phases from data in real space for each experiment. These signals are the information entropy signatures of each topological phase transition.

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for experiments 1 and 2. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1, only 6 lattice sites $S_1$ = (1, 2, 4, 51, 52, 54) corresponding to the two sharp peaks seen in figure \ref{feature_importances1} contributed more than 70\% of total reduction in Shannon entropy. Similarly, more than 30\% of total reduction in the Shannon entropy of eigenvector data from experiment 2 was achieved by 18 lattice sites $S_2$ = (1, 2, 3, 4, 5, 6, 47, 49, 50, 51, 52, 54, 95, 96, 97, 98, 99, 100) located at the three peaks in figure \ref{feature_importances2}.

@@

\begin{figure}
\centering
\subfigure[]{\label{heatmap:a}\includegraphics[width=.50\textwidth]{fig1a-6561-100-simulation_merged_winding_grid.png}}
\subfigure[]{\label{heatmap:b}\includegraphics[width=.49\textwidth]{fig1b-6561-100-simulation_merged_winding_grid.png}}
\caption{Probability heatmaps learned from real space eigenvector data. Heatmaps were averaged across all 100 iterations in both experiments. (a) The probability heatmap learned in experiment 1 shows the probability that a Hamiltonian in the grid has winding number equal to 1.  (b) The probability heatmap learned in experiment 2 is a superposition of individual probability heatmaps learned for each winding number $W$ = -1, 0, 1 and 2.}
\label{heatmaps}
\end{figure}

@@

An advantage of using data from real space is that it enables us to investigate how topological information is distributed in the system. This was demonstrated by the Gini topological signals recovered from the Gini impurity of ensembles of eigenvectors in each experiment. The existence of such signals that can be recovered from data from many distinct physical systems but are hard to conceptualize from sheer theoretical reasoning provides a clear example of how machine learning can be an important tool in the investigation and discovery of new quantum materials.

We should remark on the subtleties of the Gini topological signals. Although they give us a visualization of how important each lattice site was in determining the topological phases of Hamiltonians, these importances actually express a global property of the whole lattice. Therefore, a lattice site that appears unimportant in a Gini topological signal may not be unimportant or void of topological information by itself. To give a concrete example, reduction in Gini impurity tends to be distributed among highly correlated variables. This implies that if only a single variable in a highly correlated subset is used, it will likely inherit most of the reduction in Gini impurity from the other variables in the subset. In this regard the Gini topological signals presented here express a summary of relations between lattice sites and are therefore intrinsically global. Furthermore, it should be emphasized that these signals were recovered from the analysis of data from several thousand SSH systems in each experiment and therefore they are not a property of a single SSH lattice. They are rather a pattern that emerges from correlating topological phase with lattice eigenvector data for several SSH systems.

We performed several tests on the Gini topological signals. By rerunning each experiment with longer lattices (i.e. increasing the number of unit cells) we have verified that the Gini topological signals in figures \ref{feature_importances1} and \ref{feature_importances2} appear to converge to well defined continuous density functions in the macroscopic limit. The macroscopic limit of Gini topological signals may be an important signature of topological systems and thus merits further theoretical investigation. A detailed discussion of this point is presented in the section \textbf{Gini topological signals in the macroscopic limit} in the Supplementary Material.

Recent works have demonstrated the existence of local topological markers in real space that carry important information on the topological state of a system \cite{PhysRevB.84.241106,caio2019topological}.  Given that topological signals such as the ones shown in figures \ref{feature_importances1} and \ref{feature_importances2} are measured in terms of quantities that have actual physical meaning such as mixture entropy or Gini impurity, the results presented here suggest a new road for theoretical investigation. Whether there is any relationship between local topological markers and the entropy of mixing or the Gini impurity of the ensemble of eigenvectors is left for speculation.

@@

\begin{figure}
\centering
\subfigure[]{\label{feature_importances1}\includegraphics[width=0.49\textwidth]{fig1a-6561-100-feature_importances.png}}\quad
\subfigure[]{\label{feature_importances2}\includegraphics[width=0.49\textwidth]{fig1b-6561-100-feature_importances.png}}\quad
\caption{Information entropy signatures of the topological phase transitions studied in the paper. (a) In experiment 1, the two sharp peaks in the Shannon entropy signal account for more than 70\% of reduction in information entropy. These two peaks correspond to the lattice sites $S_1$. (b) In experiment 2, the three visible peaks account for more than 30\% of reduction in information entropy. These three peaks are located on lattice sites $S_2$.}
\label{feature_importances}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.37\textwidth]{model.pdf}
  \caption{Schematic set-up of a 1D SSH system. Here $t_1$ and $t_2$ are nearest-neighbor hoppings while $T_1$ and $T_2$ are second nearest-neighbor hoppings.}
\label{fig:model}
\end{figure}

\begin{figure}%[h!]
  \centering
  \includegraphics[width=.37\textwidth]{winding.png}
  \caption{Winding number. The winding number of a closed, oriented curve with respect to a reference point is a topological invariant that counts how many times the curve winds around the point. Picture credits: Jim Belk, public domain.}
\label{fig:winding}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{less_feat:a}\includegraphics[width=.49\textwidth]{fig1a-6561-less-simulation_merged_winding_grid.png}}
\subfigure[]{\label{less_feat:b}\includegraphics[width=.49\textwidth]{fig1b-6561-less-simulation_merged_winding_grid.png}}
\caption{Probability heatmaps learned from most relevant lattice sites. (a) Probability heatmap learned in experiment 1 using only lattice sites $S_1$.  (b) Probability heatmap learned in experiment 2 using only lattice sites $S_2$.}
\label{less_feat}
\end{figure}

@@

\subsection*{Gini topological signals in the macroscopic limit}

A natural question that comes to mind regarding the Gini topological signals presented here is whether these signals are artifacts of the numerical procedure used to generate them.

@@

There are still artifacts that might arise from the hyperparameters used to generate the data. These hyperparameters include lattice size and grid specifications (i.e., choices regarding the discretization of parameter space). Such artifacts can only be controlled for by bootstrapping over different hyperparameter settings, which may be computationally prohibitive. Intuitively, the hyperparameter we identify as likely having the most noticeable effect in the Gini topological signals is lattice size.

All results presented in the article were obtained for lattices with 50 unit cells. Each cell contains two different atoms, thus leading to 100$\times$100 Hamiltonian matrices and their corresponding eigenvectors in $\mathbb{R}^{100}$. Here we present the Gini topological signals obtained for lattices with 70, 90 and 110 unit cells for both experiments (figures \ref{newtopsig1} and \ref{newtopsig2}). These Gini topological signals were generated in exactly the same way as the Gini topological signals obtained for 50 unit cells (i.e., after bootstrapping $n_{exp}$ = 100 times and averaging lattice site relevances across all iterations).

It is clear from figures \ref{newtopsig1} and \ref{newtopsig2} that the patterns seen with 50 unit cells (figures \ref{feature_importances1} and \ref{feature_importances2}) are stable across higher lattice sizes, with finer details emerging in the signals as the lattice size increases.

Figures \ref{newtopsig1} and \ref{newtopsig2} suggest that the Gini topological signals presented here can be viewed as topological mass functions along the lattices. In the limit of an infinite chain, the cumulative Gini topological distribution function (or cumulative Gini topological distribution for short) $F_G(x)$ associated with a Gini topological signal will be given by the integral of a Gini topological density function (or Gini topological density for short) $\rho_G$($x$),

\begin{equation}
\label{entropy_dens}
F_G(x) = \int_{0}^x\rho_G(x')dx'
\end{equation}
where $x$ $\in$ [0,1] is a spatial coordinate along the lattice (being strict, our use of periodic boundary conditions implies that the coordinate $x$ should be defined in the quotient space $[0,1]/R$, where $R$ is the equivalence relation in [0,1] defining the unit circle $S^1$, i.e. $x$ $R$ $x'$ if and only if $x$ = $x'$ or ($x$,$x'$) $\in$ \{(0,1), (1,0)\}. For open boundary conditions, the spatial coordinate $x$ is defined in the closed interval [0,1]). In this continuum limit, a Gini topological signal will be given by the Gini topological density $\rho_G(x)$. Figures \ref{cdf1} and \ref{cdf2} show the cumulative Gini topological distributions corresponding to the Gini topological signals in figures \ref{newtopsig1} and \ref{newtopsig2}.

The results presented in this article suggest that the macroscopic limit $\rho_G(x)$ of Gini topological signals may have strong theoretical interest and pave new roads to the investigation of topological materials.

\begin{figure}
\centering
\subfigure[]{\label{newtopsig1-70}\includegraphics[width=.32\textwidth]{fig1a-6561-140-feature_importances}}
\subfigure[]{\label{newtopsig1-90}\includegraphics[width=.32\textwidth]{fig1a-6561-180-feature_importances}}
\subfigure[]{\label{newtopsig1-110}\includegraphics[width=.32\textwidth]{fig1a-6561-220-feature_importances}}
\caption{Gini topological signals obtained for experiment 1 with higher lattice sizes. (a) Gini topological signal for 70 unit cells. (b) Gini topological signal for 90 unit cells. (c) Gini topological signal for 110 unit cells.}
\label{newtopsig1}
\end{figure}
\begin{figure}
\centering
\subfigure[]{\label{newtopsig2-70}\includegraphics[width=.32\textwidth]{fig1b-6561-140-feature_importances}}
\subfigure[]{\label{newtopsig2-90}\includegraphics[width=.32\textwidth]{fig1b-6561-180-feature_importances}}
\subfigure[]{\label{newtopsig2-110}\includegraphics[width=.32\textwidth]{fig1b-6561-220-feature_importances}}
\caption{Gini topological signals obtained for experiment 2 with higher lattice sizes. (a) Gini topological signal for 70 unit cells. (b) Gini topological signal for 90 unit cells. (c) Gini topological signal for 110 unit cells.}
\label{newtopsig2}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{cdf1-70}\includegraphics[width=.32\textwidth]{fig1a-6561-140-cumulative_feature_importances-newer}}
\subfigure[]{\label{cdf1-90}\includegraphics[width=.32\textwidth]{fig1a-6561-180-cumulative_feature_importances-newer}}
\subfigure[]{\label{cdf1-110}\includegraphics[width=.32\textwidth]{fig1a-6561-220-cumulative_feature_importances-newer}}
\caption{Cumulative Gini topological distributions (CGTDs) for higher lattice sizes in experiment 1. (a) CGTD for 70 unit cells. (b) CGTD for 90 unit cells. (c) CGTD for 110 unit cells. The two visible leaps in the cumulative Gini topological distributions shown above occur at regions corresponding to the lattice sites $S_1$ in the case with 50 unit cells.}
\label{cdf1}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{cdf2-70}\includegraphics[width=.32\textwidth]{fig1b-6561-140-cumulative_feature_importances-newer}}
\subfigure[]{\label{cdf2-90}\includegraphics[width=.32\textwidth]{fig1b-6561-180-cumulative_feature_importances-newer}}
\subfigure[]{\label{cdf2-110}\includegraphics[width=.32\textwidth]{fig1b-6561-220-cumulative_feature_importances-newer}}
\caption{Cumulative Gini topological distributions (CGTDs) for higher lattice sizes in experiment 2. (a) CGTD for 70 unit cells. (b) CGTD for 90 unit cells. (c) CGTD for 110 unit cells. Here the leaps in the cumulative Gini topological distributions are less pronounced than in experiment 1, but still noticeable. The three visible leaps occur at regions corresponding to the lattice sites $S_2$ in the case with 50 unit cells.}
\label{cdf2}
\end{figure}

>>>>

\begin{abstract}
We develop a supervised machine learning algorithm that is able to learn topological phases for finite condensed matter systems in real lattice space. The algorithm employs diagonalization in real space together with any supervised learning algorithm to learn topological phases through an eigenvector-ensembling procedure. We combine our algorithm with decision trees to successfully recover topological phase diagrams of Su-Schrieffer-Heeger (SSH) models from bulk lattice data in real space and show how the Shannon information entropy of ensembles of lattice eigenvectors can be used to retrieve a signal detailing how topological information is distributed in the bulk. The discovery of Shannon information entropy signals associated with topological phase transitions from the analysis of data from several thousand SSH systems illustrates how model explainability in machine learning can advance the research of exotic quantum materials with properties that may power future technological applications such as qubit engineering for quantum computing. 
\end{abstract}
\begin{document}
\nocite{*}

@@

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as Shannon information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learn the desired pattern in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the whole system emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in the bulk of topological condensed matter systems is presented in the literature.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh1.png}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh2.png}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiments 1 and 2 as seen from parameter space. The accuracy statistics presented in the following subsections and probability heatmaps shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. The recovered probability heatmaps faithfully portray the phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 1 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase prediction grid learned from real space.}
\label{figexp1_exp}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 2 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase prediction grid learned from real space.}
\label{figexp2_exp}
\end{figure}

@@

The (normalized) relevance of a feature is given by how much it reduces a loss function (e.g. Shannon information entropy or Gini impurity \cite{friedman2001elements}). By averaging normalized relevances as measured by reduction in the information entropy of ensembles of real space eigenvectors across $n_{exp}$ = 100 iterations of both experiment 1 and experiment 2 we recovered Shannon entropy signals that reveal which lattice sites were consistently more relevant in learning topological phases from data in real space for each experiment. These signals are the information entropy signatures of each topological phase transition.

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for experiments 1 and 2. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1, only 6 lattice sites $S_1$ = (1, 2, 4, 51, 52, 54) corresponding to the two sharp peaks seen in figure \ref{feature_importances_ssh1} contributed more than 70\% of total reduction in Shannon entropy. Similarly, more than 30\% of total reduction in the Shannon entropy of eigenvector data from experiment 2 was achieved by 18 lattice sites $S_2$ = (1, 2, 3, 4, 5, 6, 47, 49, 50, 51, 52, 54, 95, 96, 97, 98, 99, 100) located at the three peaks in figure \ref{feature_importances_ssh2}.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh1_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh1_heatmap}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/merge_imshow_winding_grids_second_sim.png}}
\caption{Probability heatmaps learned by a combination of decision trees with eigenvector ensembling from bulk real space eigenvector data in experiment 1. Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (c) The phase diagram resulting from heatmaps (a) and (b).}
\label{ssh1_heatmaps}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{ssh2_heatmap_-1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_-1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh2_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_2}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_2_sim.png}}
\subfigure[]{\label{ssh2_heatmap}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/merge_imshow_winding_grids_sim.png}}
\caption{Probability heatmaps learned by a combination of random forests with eigenvector ensembling from bulk real space eigenvector data in experiment 2. Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to -1.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0. (c) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (d) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 2. (e) The phase diagram resulting from heatmaps (a)-(d).}
\label{ssh2_heatmaps}
\end{figure}

@@

An advantage of using data from real space is that it enables us to investigate how topological information is distributed in the system. This was demonstrated by the information entropy signatures recovered from the Shannon entropy of ensembles of eigenvectors in each experiment. The existence of such signals that can be recovered from data from many distinct physical systems but are hard to conceptualize from sheer theoretical reasoning provides a clear example of how machine learning can be an important tool in the investigation and discovery of new quantum materials.

We should remark on the subtleties of the information entropy signatures presented here. Although they give us a visualization of how important each lattice site was in determining the topological phases of Hamiltonians, these importances actually express a global property of the whole lattice. Therefore, a lattice site that appears unimportant in an information entropy signature plot may not be unimportant or void of topological information by itself. To give a concrete example, reduction in Shannon entropy tends to be distributed among highly correlated variables. This implies that if only a single lattice site in a highly correlated subset is used, it will likely inherit most of the reduction in Shannon entropy from the other lattice sites in the subset. In this regard the information entropy signatures presented here express a summary of relations between lattice sites and are therefore intrinsically global. Furthermore, it should be emphasized that these signals were recovered from the analysis of data from several thousand SSH systems in each experiment and therefore they are not a property of a single SSH lattice. They are rather a pattern that emerges from correlating topological phase with lattice eigenvector data for several SSH systems.

We performed several tests on the information entropy signatures. By rerunning each experiment with longer lattices (i.e. increasing the number of unit cells) we have verified that the signals in figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2} appear to converge to well defined continuous density functions in the macroscopic limit. The macroscopic limit of these information entropy signals may be an important signature of topological systems and thus merits further theoretical investigation. A detailed discussion of this point is presented in the section \textbf{Information entropy signatures in the macroscopic limit} in the Supplementary Material.

Recent works have demonstrated the existence of local topological markers in real space that carry important information on the topological state of a system \cite{PhysRevB.84.241106,caio2019topological}.  Given that topological signals such as the ones shown in figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2} are measured in terms of quantities that have actual physical meaning such as information entropy or Gini impurity, the results presented here suggest a new road for theoretical investigation. Whether there is any relationship between local topological markers and the information entropy of the ensemble of eigenvectors is left for speculation.

@@

\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh1}\includegraphics[width=0.49\textwidth]{./ssh1/periodic_100_6561/plot_feature_importances.png}}\quad
\subfigure[]{\label{feature_importances_ssh2}\includegraphics[width=0.49\textwidth]{./ssh2/periodic_100_6561/plot_feature_importances.png}}\quad
\caption{Information entropy signatures of the topological phase transitions from experiments 1 and 2. (a) In experiment 1, the two sharp peaks in the Shannon entropy signal account for more than 70\% of reduction in information entropy. These two peaks correspond to the lattice sites $S_1$. (b) In experiment 2, the three visible peaks account for more than 30\% of reduction in information entropy. These three peaks are located on lattice sites $S_2$.}
\label{feature_importances}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.37\textwidth]{./phase_diagrams/ssh_model.pdf}
  \caption{Schematic set-up of a 1D SSH system. Here $t_1$ and $t_2$ are nearest-neighbor hoppings while $T_1$ and $T_2$ are second nearest-neighbor hoppings.}
\label{fig:model}
\end{figure}

\begin{figure}%[h!]
  \centering
  \includegraphics[width=.37\textwidth]{./phase_diagrams/winding.png}
  \caption{Winding number. The winding number of a closed, oriented curve with respect to a reference point is a topological invariant that counts how many times the curve winds around the point. Picture credits: Jim Belk, public domain.}
\label{fig:winding}
\end{figure}

@@

%\begin{figure}
%\centering
%\subfigure[]{\label{less_feat:a}\includegraphics[width=.49\textwidth]{fig1a-6561-less-simulation_merged_winding_grid.png}}
%\subfigure[]{\label{less_feat:b}\includegraphics[width=.49\textwidth]{fig1b-6561-less-simulation_merged_winding_grid.png}}
%\caption{Probability heatmaps learned from most relevant lattice sites. (a) Probability heatmap learned in experiment 1 using only lattice %sites $S_1$.  (b) Probability heatmap learned in experiment 2 using only lattice sites $S_2$.}
%\label{less_feat}
%\end{figure}

@@

\subsection*{Information entropy signatures in the macroscopic limit}

A natural question that comes to mind regarding the information entropy signatures presented here is whether these signals are artifacts of the numerical procedure used to generate them.

@@

There are still artifacts that might arise from the hyperparameters used to generate the data. These hyperparameters include lattice size and grid specifications (i.e., choices regarding the discretization of parameter space). Such artifacts can only be controlled for by bootstrapping over different hyperparameter settings, which may be computationally prohibitive. Intuitively, the hyperparameter we identify as likely having the most noticeable effect in the information entropy signatures is lattice size.

All results presented in the article were obtained for lattices with 50 unit cells. Each cell contains two different atoms, thus leading to 100$\times$100 Hamiltonian matrices and their corresponding eigenvectors in $\mathbb{R}^{100}$. Here we present the information entropy signatures obtained for lattices with 70, 90 and 110 unit cells for both experiments (figures \ref{newtopsig1} and \ref{newtopsig2}). These signals were generated in exactly the same way as the information entropy signatures obtained for 50 unit cells (i.e., after bootstrapping $n_{exp}$ = 100 times and averaging lattice site relevances across all iterations).

It is clear from figures \ref{newtopsig1} and \ref{newtopsig2} that the patterns seen with 50 unit cells (figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2}) are stable across higher lattice sizes, with finer details emerging in the signals as the lattice size increases.

Figures \ref{newtopsig1} and \ref{newtopsig2} suggest that the information entropy signatures presented here can be viewed as entropy mass functions along the lattices. In the limit of an infinite chain, the cumulative entropy distribution $F_S(x)$ associated with a Shannon information entropy signal will be given by the integral of an entropy density function $\rho_S$($x$),

\begin{equation}
\label{entropy_dens}
F_S(x) = \int_{0}^x\rho_S(x')dx'
\end{equation}
where $x$ $\in$ [0,1] is a spatial coordinate along the lattice (being strict, our use of periodic boundary conditions implies that the coordinate $x$ should be defined in the quotient space $[0,1]/R$, where $R$ is the equivalence relation in [0,1] defining the unit circle $S^1$, i.e. $x$ $R$ $x'$ if and only if $x$ = $x'$ or ($x$,$x'$) $\in$ \{(0,1), (1,0)\}. For open boundary conditions, the spatial coordinate $x$ is defined in the closed interval [0,1]). In this continuum limit, an information entropy signature will be given by the entropy density function $\rho_S(x)$. Figures \ref{cdf1} and \ref{cdf2} show the cumulative entropy distributions corresponding to the information entropy signatures in figures \ref{newtopsig1} and \ref{newtopsig2}.

The results presented in this article suggest that the macroscopic limit $\rho_S(x)$ of the information entropy signatures may have strong theoretical interest and pave new roads to the investigation of topological materials.

\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh1_140}\includegraphics[width=.32\textwidth]{./ssh1/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_180}\includegraphics[width=.32\textwidth]{./ssh1/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_220}\includegraphics[width=.32\textwidth]{./ssh1/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signatures obtained for experiment 1 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh1_longer_lattices}
\end{figure}
\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signature obtained for experiment 2 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh2_longer_lattices}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{cumulative_entropy_ssh1_140}\includegraphics[width=.32\textwidth]{./ssh1/periodic_140_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh1_180}\includegraphics[width=.32\textwidth]{./ssh1/periodic_180_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh1_220}\includegraphics[width=.32\textwidth]{./ssh1/periodic_220_6561/plot_cumulative_feature_importances}}
\caption{Cumulative entropy distributions for higher lattice sizes in experiment 1. (a) Cumulative entropy distribution for 70 unit cells. (b) Cumulative entropy distribution for 90 unit cells. (c) Cumulative entropy distribution for 110 unit cells. The two visible leaps in the cumulative entropy distributions shown above occur at regions corresponding to the lattice sites $S_1$ in the case with 50 unit cells.}
\label{cdf1}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{cumulative_entropy_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_cumulative_feature_importances}}
\caption{Cumulative entropy distributions distributions for higher lattice sizes in experiment 2. (a) Cumulative entropy distributions for 70 unit cells. (b) Cumulative entropy distributions for 90 unit cells. (c) Cumulative entropy distributions for 110 unit cells. Here the leaps in the cumulative entropy distributions are less pronounced than in experiment 1, but still noticeable. The three visible leaps occur at regions corresponding to the lattice sites $S_2$ in the case with 50 unit cells.}
\label{cdf2}
\end{figure}
